Confirmatory factor analysis (CFA) of multivariate lipid distributions across

### 2017-09-07 draft/ms-201705-draft.Rmd
update to draft

---
title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups -- United States, 2003 to 2012
author: "Ann Von Holle, Kari E. North"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    csl: ama.csl
bibliography: bib1.bib
csl: ama.csl
---

@@

When comparing gender, racial/ethnic, and age groups, the metric model, allowing intercepts to vary but fixed factor loadings, was not significantly different (Chi-Square=5.6, 2 df, p<0.06; Chi-Square=11.6, 8 df, p>0.16; Chi-Square=8.8, 4 df, p>0.06, respectively) from the configural model, which allowed both loading and intercepts to vary, indicating similarity of factor loadings. On the other hand, the scalar model, fixing both intercepts and factor loadings was different from the configural model (Chi-Square=71.1, 5 df, p<0.0001; Chi-Square=189.2, 20 df, p<0.0001; Chi-Square=86.5, 10 df, p<0.0001, respectively), indicating a difference in factor intercepts. Evidence for different factor intercepts is further supported by a significant test of the scalar versus metric model (Chi-Square=68.8, 3 df, p<0.0001; Chi-Square=198.8, 12 df, p<0.0001, Chi-Square=82.0, 6 df, p<0.0001, respectively). Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts -- the metric model. [Due to these results it is worth noting that a strict interpretation of factor means across the groups is not possible given the different intercepts. Add this part to discussion below]

Model fit information (Table 1) was satisfactory for the analyses by gender, racial/ethnic and age groups. First, RMSEA was below 0.04 in all three cases. Second, CFI was above 0.98. Third, aBIC was lowest for the metric model for all three models with the exception of the gender groups, and in this scenario, the difference between the configural and metric model was less than 4 points, considered to be a negligible difference. All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters.

@@

Finally, factor loadings are not allowed to vary across groups in the metric model. Factor loadings indicate the degree to which the measured correlation between the indicators are correlated with the dyslipidemia factor. TG has the strongest correlation with the factor, at a correlation exceeding 0.71. HDL has the second strongest association -- this time a negative one -- around -0.52. Finally, LDL is the least correlated with the factor with a correlation at 0.39 for this sample. Any standardized factor loading below 0.3 indicates need to remove that indicator, and none of the lipid indicators fit that description. These values culminate in a factor indicating that a one unit increase in the factor translates to adverse levels of lipids with a decline in HDL and increases in TG and LDL.

@@

# Conclusion

# Tables

```{r chi, child='../mplus-results-manuscript.Rmd'}

```

>>>>

---
title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across
  time and race-ethnic groups -- United States, 2003 to 2012
author: "Ann Von Holle, Kari E. North"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    csl: ama.csl
    depth: 3
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
  word_document:
    toc: yes
csl: ama.csl
bibliography: bib1.bib
---
---
title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003-2012
author: "Ann Von Holle, Kari E. North"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    csl: ama.csl
bibliography: bib1.bib
csl: ama.csl
---

@@

When comparing gender, racial/ethnic, and age groups, the metric model, allowing intercepts to vary but fixed factor loadings, was not significantly different (Chi-Square=5.6, 2 df, p<0.06; Chi-Square=11.6, 8 df, p>0.16; Chi-Square=8.8, 4 df, p>0.06, respectively) from the configural model, which allowed both loading and intercepts to vary, indicating similarity of factor loadings. On the other hand, the scalar model, fixing both intercepts and factor loadings was different from the configural model (Chi-Square=71.1, 5 df, p<0.0001; Chi-Square=189.2, 20 df, p<0.0001; Chi-Square=86.5, 10 df, p<0.0001, respectively), indicating a difference in factor intercepts. Evidence for different factor intercepts is further supported by a significant test of the scalar versus metric model (Chi-Square=68.8, 3 df, p<0.0001; Chi-Square=198.8, 12 df, p<0.0001, Chi-Square=82.0, 6 df, p<0.0001, respectively). Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts -- the metric model.

Model fit information (Table 1) was satisfactory for the analyses by gender, racial/ethnic and age groups. First, RMSEA was below 0.05 in all three cases. Second, CFI was above 0.98. Third, aBIC was lowest for the metric model for all three models with the exception of the gender groups, and in this scenario, the difference between the configural and metric model was less than 4 points, considered to be a negligible difference. All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters [@brown_confirmatory_2015].

@@

Finally, factor loadings are not allowed to vary across groups in the metric model. Factor loadings indicate the degree to which the measured correlation between the indicators are correlated with the dyslipidemia factor concurrently separating out the measurement error -- an advantage of CFA. Examining these association shows that TG has the strongest correlation with the factor, at a correlation exceeding 0.71. HDL has the second strongest association -- this time a negative one -- around -0.52. Finally, LDL is the least correlated with the factor with a correlation at 0.39 for this sample. Any standardized factor loading below 0.3 indicates need to remove that indicator, and none of the lipid indicators fit that description. These values culminate in an interpretation of a one unit increase in the factor equating with an increase in adverse levels of lipids: declining HDL and increasing TG and LDL.

@@

The purpose of this paper was to examine the use of CFA as a measure of dyslipidemia. First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG. Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- a common area of comparisons for lipid measures. Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept. MI did not hold for different racial/ethnic nor to age groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.

## Measurement Invariance

In evaluating the three different types of models across subgroups for each of the following domains: age, race/ethnicity and sex of child, the metric model was the best of the three for all domains. The configural model, in which all factor loadings and means of the dyslipidemia factor were allowed to vary across groups, was not statistically significantly different from the metric model. Given this result, the metric model offers a parsimonious fit by fixing the factor loadings to be the same across groups. Furthermore, the metric model was different from the scalar model, adding more evidence to support the choice of the metric model.

Similar factor loadings across the subgroups implies correlations between the three lipid indicators are the same across the subgroups within each domain. This finding may have relevance for lipid ratio measures such as HDL to TG, a commonly used metric. For example, a higher factor score for a construct with positive HDL-C factor loadings combined with negative LDL-C and TG factor loadings would mean on average a simultaneous increase of HDL and decrease in LDL-C and TG, equivalent to a higher HDL-C to LDL-C ratio. This interpretation implies higher ratios are analogous to higher factor score as well as lower ratios having lower factor scores.

Although the factor loadings for the construct are similar, this finding does not imply metric invariance holds for these subgroups. In fact, the measurement invariance testing does not support comparing factor means across subgroups within each domain. Metric invariance implies the centers of this multidimensional construct are different across the subgroups. For example, when the factor is equal to zero for all subgroups, the estimated indicator (HDL-C, LDL-C and TG) are all different depending on the subgroup you choose. This result has implications regarding use of mean factor scores across subgroups. Some researchers tends to obtain factor scores and use them as observed entities, then comparing them across subgroups. These analyses demonstrate the perils in this approach with scientifically meaningless results given the evidence of scale differences. The existence of measurement non-invariance supports different dyslipidemia constructs for each of the selected subgroups for race/ethnicity, age and gender. As such, comparisons in factor score means have different meanings and are not substantively meaningful. Although this finding remains restricted to this example, it pertains to a nationwide representative sample of children in the United States and provides an important reminder when conducting analyses in other samples.

## Fit

Finally, having chosen the metric model for all of the three domains, we determined that model fit diagnostics were within normal bounds of adequate to good fit and as such motivates further use of this construct. Use of this construct would be an interesting supplement to univariate outcomes with a different and similarly meaningful interpretation.

Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research. Lipids do not function independently of one another and a measure addressing their covariance is an asset. CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators. Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.

This alternate construct can be an adjunct measure to univariate measures of lipids, which provides a means to assess the degree to which one measure increases as another measure either increase or decreases on a population level. Aside from the important caveat of measurement non-invariance, this work demonstrates the degree to which HDL-C, LDL-C and TG are correlated with one another on a population level. In turn this correlation can provide meaningful associations with other factors in downstream analyses.

# References
